Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-20 manuscript.tex
Edits to Discussion cont.

Significant effects of the \emph{Hopscotch} insertion on lateral branch length, number of cupules, and tillering index in domesticated maize have been well documented \citep{Studer2011}.  \citet{Weber2007} have described significant phenotypic associations between markers in and around \emph{tb1} and lateral branch length and female ear length within a sample from 74 natural populations of \emph{parviglumis} \citep{Weber2007}; these data did not include markers from the \emph{Hopscotch} region 66kb upstream of \emph{tb1}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion across a wide collection of individuals sampled from natural populations of teosinte. We have found no significant effect of the \emph{Hopscotch} insertion on tillering index or tiller number, a result that is discordant with its clear phenotypic effects in maize. One interpretation of this result would be that the \emph{Hopscotch} controls tillering in maize \citep{Studer2011}, but tillering in teosinte is affected by variation at other loci. In fact, \emph{tb1} is thought to be part of a complex pathway controlling branching, tillering and other phenotypic traits \citep{KebromBrutnell2007, Clark2006}. A recent study by \citet{StuderDoebley2012} examined variation across traits in a three-taxa allelic series at the \emph{tb1} locus. \citet{StuderDoebley2012} introgressed nine unique teosinte \emph{tb1} segments (one from \emph{Zea diploperennis}, and four each from \emph{mexicana} and \emph{parviglumis}) into an inbred maize background and investigated phenotypic effects. Phenotypes were shown to cluster by taxon, indicating \emph{tb1} potentially played a role in the morphological diversification of \emph{Zea}. Additional analysis suggested tillering index was controlled both by \emph{tb1} and loci elsewhere in the genome.  Clues to the identity of these loci may be found in QTL studies that identified loci controlling branching architecture \citep{DoebleyStec1991, DoebleyStec1993}. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012},  and both \emph{tru1} and \emph{ter1} affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number; \citealt{Li2012}). Variation in these additional loci may have affected tillering in our collections and contributed to the lack of correlation we see between \emph{Hopscotch} genotype and tillering.

In summary, our findings demonstrate that the \emph{Hopscotch} allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression; however, it seems unlikely that the insertion would have drifted to high frequency in multiple populations. The \emph{Hopscotch} does not appear to reduce tillering in \emph{parviglumis} as it does in maize. Other loci involved in branching architecture may regulate tillering in teosinte. Finally, we find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}; this coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations suggests that the locus plays an ecological role in teosinte. In the future, additional experiments will be needed to examine expression levels of \emph{tb1} and additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{ter1}) in conjunction with a more exhaustive phenotyping and genotyping assay. \jri{why not Phyb and phya? } \lev{Are they necessary to include? I'd had them in before in a paragraph but had been voted out} \jri{I'd ditch gt1 tru1 ter1 and maybe just cite some people including phyb etc.}

>>>>

Significant effects of the \emph{Hopscotch} insertion on lateral branch length, number of cupules, and tillering index in domesticated maize have been well documented \citep{Studer2011}.  \citet{Weber2007} have described significant phenotypic associations between markers in and around \emph{tb1} and lateral branch length and female ear length in a sample from 74 natural populations of \emph{parviglumis} \citep{Weber2007}; however, these data did not include markers from the \emph{Hopscotch} region 66kb upstream of \emph{tb1}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion across a wide collection of individuals sampled from natural populations of teosinte. We have found no significant effect of the \emph{Hopscotch} insertion on tillering index or tiller number, a result that is discordant with its clear phenotypic effects in maize. One interpretation of this result would be that the \emph{Hopscotch} controls tillering in maize \citep{Studer2011}, but tillering in teosinte is affected by variation at other loci. Consistent with this interpretation, \emph{tb1} is thought to be part of a complex pathway controlling branching, tillering and other phenotypic traits \citep{KebromBrutnell2007, Clark2006}. A recent study by \citet{StuderDoebley2012} examined variation across traits in a three-taxa allelic series at the \emph{tb1} locus. \citet{StuderDoebley2012} introgressed nine unique teosinte \emph{tb1} segments (one from \emph{Zea diploperennis}, and four each from \emph{mexicana} and \emph{parviglumis}) into an inbred maize background and investigated phenotypic effects. Phenotypes were shown to cluster by taxon, indicating \emph{tb1} may underlie morphological diversification of \emph{Zea}. Additional analysis in \citet{StuderDoebley2012} suggested tillering index was controlled both by \emph{tb1} and loci elsewhere in the genome.  Clues to the identity of these loci may be found in QTL studies that have identified loci controlling branching architecture (\emph{e.g.,} \citealt{DoebleyStec1991, DoebleyStec1993}). Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012},  and both \emph{tru1} and \emph{ter1} affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number; \citealt{Li2012}). Variation in these additional loci may have affected tillering in our collections and contributed to the lack of correlation we see between \emph{Hopscotch} genotype and tillering.

In conclusion, our findings demonstrate that the \emph{Hopscotch} allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression; however, it seems unlikely that the insertion would have drifted to high frequency in multiple populations. We do, however, find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}; this coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations suggests that the locus plays an ecological role in teosinte. In contrast to domesticated maize, the \emph{Hopscotch} insertion in \emph{parviglumis} does not appear to reduce tillering. Other loci involved in branching architecture may regulate tillering in teosinte. Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{ter1}).  These data, in conjunction with more exhaustive phenotyping, should help reveal the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte. \jri{why not Phyb and phya? } \lev{Are they necessary to include? I'd had them in before in a paragraph but had been voted out} \jri{I'd ditch gt1 tru1 ter1 and maybe just cite some people including phyb etc.}
